Certified Symbolic Transducer with Applications in String Solving

### 2025-03-29 Paper/main.tex
rerun test
update code

\documentclass[a4paper,UKenglish,cleveref, autoref, anonymous, thm-restate]{lipics-v2021}
%This is a template for producing LIPIcs articles. 
%See lipics-v2021-authors-guidelines.pdf for further information.
%for A4 paper format use option "a4paper", for US-letter use option "letterpaper"
%for british hyphenation rules use option "UKenglish", for american hyphenation rules use option "USenglish"
%for section-numbered lemmas etc., use "numberwithinsect"
%for enabling cleveref support, use "cleveref"
%for enabling autoref support, use "autoref"
%for anonymousing the authors (e.g. for double-blind review), add "anonymous"
%for enabling thm-restate support, use "thm-restate"
%for enabling a two-column layout for the author/affilation part (only applicable for > 6 authors), use "authorcolumns"
%for producing a PDF according the PDF/A standard, add "pdfa"

@@

%TODO mandatory: add short abstract of the document
\begin{abstract}
Finite-state Automata (FAs) are fundamental components in the domains of programming languages. For instance, regular expressions, which are pivotal in languages such as JavaScript and Python, are frequently implemented using FAs.
%
Finite-state Transducers (FTs) extend the capabilities of FAs by enabling the transformation of input strings into output strings, thereby providing a more expressive framework for operations that encompass both recognition and transformation. Despite the widespread adoption of formalizing FAs and FTs within proof assistants such as Coq and Isabelle/HOL, these formalizations often fall short in terms of applicability to real-world scenarios. A significant limitation of classical finite-state models is that transition labels are typically defined as a single character from a finite alphabet. However, in practical applications, the alphabet of an FA or FT can be extensive or even \emph{infinite}. This classical approach to formalizing transitions can result in transition explosion, leading to critical performance bottlenecks of FA and FT operations.

@@

Finite-state Automata (FA) and Finite-state Transducers (FT) are fundamental constructs in the theory of formal languages, with extensive applications in programming languages and software engineering. For example, recent advancements in string solvers, as demonstrated by \cite{pacmpl/ChenFHHHKLRW22}, have illustrated the relationship between regular expressions in modern programming languages and various forms of FAs and FTs. Additionally, FAs and FTs find significant industrial applications, such as in the verification of AWS access control policies \cite{DBLP:conf/fmcad/BackesBCDGLRTV18}.

@@

Symbolic Finite-state Automata (SFA) and Symbolic Finite-state Transducers (SFT) \cite{cav/DAntoniV17, VeanesHLMB12Transducer} represent advanced extensions of classic FAs and  FTs, enhancing their applicability in practical scenarios.
%
These symbolic models utilize transition labels defined by boolean algebras, allowing for more expressive representations. For example, a transition label might be specified as an interval ($'\texttt{a}'-'\texttt{z}'$), encompassing all characters from $'\texttt{a}'$ to $'\texttt{z}'$, or as an arithmetic condition ($x \mod 2 = 0$), denoting all even numbers. This symbolic approach not only provides a more succinct representation but also supports infinite alphabets, thereby extending the expressive capabilities of FAs and FTs.

The formalization of transition labels in SFAs and SFTs presents significant challenges within interactive proof assistants. Two fundamental considerations arise: first, the representation of transition labels must be sufficiently expressive to accommodate diverse boolean algebras; second, the formalization framework must be designed with extensibility in mind to facilitate the incorporation of new boolean algebras while minimizing redundant proof efforts.

Prior work of CertiStr \cite{cpp/KanLRS22} has successfully formalized SFAs within Isabelle/HOL, demonstrating both the efficiency and effectiveness of SFAs in practice. However, the formalization of FTs remains an open challenge, primarily due to their inherent complexity in two aspects: the formalization of transition labels and the specification of transition output functions. FTs constitute a significantly more expressive and powerful theoretical framework compared to symbolic FAs, as evidenced by their capability to model complex string transformations such as replacement operations.
%
Furthermore, numerous studies have demonstrated the broad applicability of SFTs across diverse domains. The work of \cite{VeanesHLMB12Transducer} showcases SFTs in security-critical applications, particularly for cross-site scripting (XSS) prevention. \cite{uss/HooimeijerLMSV11} extends this security focus by applying SFTs to web application sanitizer analysis. In system verification, \cite{osdi/YaseenABCL20} employs SFTs for runtime behavior monitoring, while \cite{pldi/HuD17} demonstrates their utility in automated program transformation through systematic inversion techniques.

In this work, we present a comprehensive formalization of SFTs. To address the extensibility challenges inherent in supporting diverse transition label theories, we adopt a refinement-based approach. At the abstraction level, transition labels are formalized through the fundamental mathematical concept of \emph{sets}. This abstraction facilitates subsequent refinement to various boolean algebras, including intervals and arithmetic conditions, while maintaining theoretical consistency.

The key operation of our formalization is the product operation between an SFT and an input regular language. Specifically, given an SFA $\mathcal{A}$ representing a regular language and an SFT $\mathcal{T}$, we define the product operation $\mathcal{T} \times\mathcal{A}$ that characterizes the set of output languages generated by $\mathcal{T}$ when processing inputs from the language recognized by $\mathcal{A}$.

In the refinement level, we implemented transition labels using an interval-based representation. The formalization of interval algebra provides efficient set-theoretic operations—including membership checking, intersection, and difference computations—facilitating the refinement of transition labels from abstract sets to concrete intervals. Furthermore, leveraging the data refinement framework \cite{DBLP:conf/itp/Lammich13} in Isabelle/HOL, we store states and transitions using sophisticated data structures such as hashmaps and red-black trees, ensuring efficient automata manipulation.

@@

In summary, our formalization makes the following contributions:
\begin{enumerate}
\item We present the first formalization of symbolic finite-state transducers in the Isabelle/HOL proof assistant.
\item We leverage the refinement framework to create an extensible SFT formalization that is capable of accommodates various boolean algebras, ensuring adaptability to future transition label developments.
\item We develop an interval algebra with efficient set-theoretic operations, enabling refinement of transition labels from abstract sets to concrete intervals.
\item We demonstrate the practical utility of our formalization through its application to string constraint solving, specifically in modeling replacement operations with verified correctness guarantees.
\end{enumerate}

The remainder of this paper is organized as follows:
Section \ref{sec:formalization} introduces the formal foundations of symbolic finite-state transducers.
Section \ref{sec:product-operation} presents the product operation at the abstract level.
Section \ref{sec_alg_refinement} details the algorithmic refinement of the product operation.
Section \ref{sec-app-str-solver} demonstrates the application to string constraint solving.
Section \ref{sec:related-work} discusses related work.
Section \ref{sec:conclusion} concludes with future directions.

\section{Formalization of Symbolic FTs}
\label{sec:formalization}

We begin by presenting a mathematical definition of SFTs \cite{VeanesHLMB12Transducer}, abstracting from the specific implementation details of Isabelle/HOL. The foundational concept underlying this formalization is the label theory, which characterizes the process by which output labels are derived from input labels.

@@

\begin{definition}[Symbolic Finite-state Transducer]
\label{def-sft}
   A Symbolic Finite-state Transducer over $\tau_1\rightarrow \tau_2$ is a quadruple $\mathcal{T} = (\mathcal{Q}, \Delta, \mathcal{I}, \mathcal{F})$, where 
   \begin{itemize}
   \item $\mathcal{Q}$ is a finite set of states,
   \item $\mathcal{I}\subseteq \mathcal{Q}$ is the set of initial states,
   \item $\mathcal{F} \subseteq\mathcal{Q}$ is the set of accepting states,
   \item $\Delta$ is the set of transition relations. Each element in $\Delta$ is of the form $(q, \phi, f, q')$ or written as $q\xrightarrow{\phi, f} q'$, where $q$ and $q'$ are states in $\mathcal{Q}$.
   $\phi$ is a predicate of type $\tau_1\rightarrow \mathbb{B}$.
   $f$ is a lambda term of type $\tau_1\rightarrow \tau_2$. $f$ is called an \emph{output function}.
   \end{itemize}

@@

    Figure \ref{fig-example-ft} illustrates an SFT that accepts strings matching the regular expression $/['\texttt{a}'-'\texttt{z}']+['\texttt{A}'-'\texttt{Z}']+/$. The transition labels utilize intervals of the form $[i\text{-}j]$, which are interpreted as predicates $\lambda x.~i \leq x \leq j$. The output functions $f_1$ and $f_2$ perform case transformations: $f_1 = \lambda x.~\texttt{toUpper}(x)$ converts lowercase letters to uppercase, while $f_2 = \lambda x.~\texttt{toLower}(x)$ performs the inverse operation. For example, given the input string \texttt{"bigSMALL"}, this SFT produces the output \texttt{"BIGsmall"}.

@@

\begin{figure}[hbt!]
	\begin{lstlisting}
record (|$'q,$||$~'a,$| |$'b$|) |$\texttt{NFT}$| =
	|$\mathcal{Q}_t$| :: "|$'q$| set"
	|$\Delta_t$| :: "(|$'q,$||$~'a,$| |$'b$|) LTTS"
	|$\mathcal{I}_t$| :: "|$'q$| set"
	|$\mathcal{F}_t$| :: "|$'q$| set"
        |$\mathcal{M}_t$| :: "|$'b\Rightarrow 'a$| Tlabel"

type_synonym |$('q, 'a, 'b)$| LTTS = "|$('q \times ('a \text{ set option }\times b) \times 'q)$| set"

type_synonym |$'a$| Tlabel = "|$'a \text{ option}~\Rightarrow~'b$ set option"
	\end{lstlisting}
\caption{The formalization of $FTs$ in Isabelle/HOL}
\label{fig-def-FT}
\end{figure}

Figure \ref{fig-def-FT} presents our formalization of SFTs in Isabelle/HOL. While the elements $\mathcal{Q}_t$, $\mathcal{I}_t$, and $\mathcal{F}_t$ directly correspond to their counterparts ($\mathcal{Q}$, $\mathcal{I}$, and $\mathcal{F}$) in Definition \ref{def-sft}, the transition relations are not exactly the same.
%
The transition relations are formalized through \texttt{LTTS} (Labeled Transducer Transition System), where each transition is represented as a triple $'q \times ('a \text{ set option }\times b) \times 'q$. This representation reflects several key design decisions aimed at enhancing the abstraction and flexibility of our SFT formalization.

Firstly, $'a \text{ set option }$ is the input type of the transition, it accepts a set of elements of type $'a$ or $\texttt{None}$ corresponding to empty string $\varepsilon$. 
Accepting a set of $'a$ elements aims to express the same but more abstract semantics of the input labels in Definition \ref{def-sft}, in which an input label is a predicate. A predicate's semantics as introduced before represents a set of elements that make the predicate true. But predicates have various different forms. For instance, the interval $[1, 9]$ represents the set $\{e \mid 1 \leq e \leq 9\}$. The predicate $\lambda b.~ b[7] = 1$, where $b$ is a bit vector of length 8, denotes the set of bit vectors that have a 1 in the 7th position. All these different forms of labels are abstracted as sets in our formalization.
%
The value $\texttt{None}$ represents the empty string $\varepsilon$ in our formalization, indicating a transition that consumes no input but may still produce output elements. This design choice facilitates the modeling of real-world applications in SFTs, as we will demonstrate in our application to string solvers.

The second element of type $'b$ serves as an index into the output function space. The mapping $\mathcal{M}_t$ associates each index with a specific output function. These output functions, formalized by \texttt{Tlabel}, map a single input element to a set of possible output elements rather than to a single element. This design enables non-deterministic output behavior, where the transducer may select any element from the output set randomly or according to specified criteria. Additionally, output functions can produce the empty string $\varepsilon$ by returning $\texttt{None}$, providing further flexibility in transition behavior.

@@

\section{The Product Operation of NFTs}
\label{sec:product-operation}

In this section, we formalize the product operation between an SFT $\mathcal{T}$ and an SFA $\mathcal{A}$, denoted as $\mathcal{T} \times \mathcal{A}$ in Section \ref{sec:formalization}. An additional consideration in this operation is the presence of $\varepsilon$-transitions in our SFT formalization, which implies that the resulting automata may also contain $\varepsilon$-transitions. Since CertiStr \cite{cpp/KanLRS22} does not include a formalization of automata with $\varepsilon$-transitions ($\varepsilon$NFAs), we extend the framework with a formalization of symbolic $\varepsilon$NFAs (denoted as $\varepsilon$SFA) and provide a verified conversion to standard SFAs. In this paper, we present only the definitions of  $\varepsilon$SFAs and SFAs, while the complete formalization, including correctness proofs, is available in our Isabelle development.

@@

type_synonym |$('q, 'a)$| LTS = "|$'q \times 'a \text{ set }\times 'q$|"    
	\end{lstlisting}
\caption{The formalization of $\varepsilon$NFA and SFA in Isabelle/HOL}
\label{fig-def-FAs}
\end{figure}

@@

\begin{figure}[hbt!]
	\begin{lstlisting}
definition productT :: |$"('q,'a,'b,'c)"$| NFT |$\Rightarrow$| ('q,'a) NFA |$\Rightarrow$| 
           |$(('a,'b)$| Tlabel |$\Rightarrow 'a \text{ set}\Rightarrow 'b\text{ set option})\Rightarrow$|
           |$('q\times 'q, ~'b)$| eNFA where
  "productT |$\mathcal{T}$| |$\mathcal{A}$| F = |$\llparenthesis$|
    |$\mathcal{Q}e = \mathcal{Q}_t~\mathcal{T} \times \mathcal{Q} ~\mathcal{A}$|,
    |$\Delta_e = \{((p,p'),~\texttt{the }(((\mathcal{M}~\mathcal{T})~f)\texttt{ None}),~(q, p'))\mid p, p', q, f.~p'\in \mathcal{Q}~\mathcal{A} \land$|
          |$(p, (\texttt{None}, f), q)\in \Delta_T \mathcal{T} \land \exists S. (\mathcal{M}~\mathcal{T})~f~\text{None} = \text{Some } S\} ~\cup$|
        |$\{((p,p'),~\texttt{the }(F~((\mathcal{M}~\mathcal{T})~f)~(\sigma_1\cap\sigma_2)),~(q, q'))\mid p, p', q, \sigma_1, \sigma_2, q', f.~$|
          |$ (p, (\text{Some }\sigma_1, f), q)\in \Delta_T~\mathcal{T} \land (p', \sigma_2, q')\in\Delta~\mathcal{A}\land \sigma_1\cap \sigma_2 = \emptyset \land$|
            |$ \exists S.~F~((\mathcal{M}~\mathcal{T})~f)~(\sigma_1\cap\sigma_2) = \text{Some } S\}$|,
    |$\Delta_e' = \{((p,p'),~\text{the }(((\mathcal{M}~\mathcal{T})~f)\texttt{ None}),~(q, p'))\mid p, p', q, f.~p'\in \mathcal{Q}~\mathcal{A} \land$|
          |$(p, (None, f), q)\in \Delta_T~\mathcal{T} \land (\mathcal{M}~\mathcal{T})~f~\texttt{None} = \text{None}\} ~\cup$|
        |$\{((p,p'),~\texttt{the }(F~((\mathcal{M}~\mathcal{T})~f)~(\sigma_1\cap\sigma_2)),~(q, q'))\mid p, p', q, \sigma_1, \sigma_2, q', f.~$|
          |$ (p, (\text{Some }\sigma_1, f), q)\in \Delta_T~\mathcal{T} \land (p', \sigma_2, q')\in\Delta~\mathcal{A}\land \sigma_1\cap \sigma_2 = \emptyset \land$|
            |$ \exists x\in(\sigma_1\cap\sigma_2).~((\mathcal{M}~\mathcal{T})~f)~(\text{Some } x) = \text{None} \}$|,
    |$\mathcal{I}_e = \mathcal{I}_t~\mathcal{T} \times \mathcal{I}~\mathcal{A}$|
    |$\mathcal{F}_e = \mathcal{F}_t~\mathcal{T} \times \mathcal{F}~\mathcal{A}$| |$\rrparenthesis$|"
	\end{lstlisting}
\caption{The formalization of $\varepsilon-$NFA and NFA in Isabelle/HOL}
\label{fig-def-FTProd}
\end{figure}

Figure \ref{fig-def-FTProd} depicts the abstract level formalization for the product of an SFT and an SFA. The parameters $\mathcal{T}$ and $\mathcal{A}$ are an SFT and an SFA, respectively. The 
result of the product operation is an $\varepsilon$SFA.
But we need to explain the role of $\text{\texttt{F}}$. The output function $f$ for each transition in $\Delta_t$ is of type "$'a\;\text{option} \Rightarrow 'b\;\text{set option}$", which applies to a \emph{single} element of type $'a$ or $\varepsilon$. $\text{\texttt{F}}$ extends $f$ to apply it to a set of elements. More precisely, let $f$ be an output function, the semantics of $\text{\texttt{F}}$ is defined as

\[\texttt{F}~f~A=\bigcup_{a\in A} (\text{if }f~a= \text{Some }S \texttt{ then } S \texttt{ else } \emptyset)\].

The transition relations $\Delta_e$ and $\Delta_e'$ are determinted by considering two distinct cases based on the nature of transitions in the SFT: $\varepsilon$-transitions and non-$\varepsilon$-transitions. In both cases, the transition labels in the resulting $\varepsilon$SFA are derived from the composition of the SFT's output function and the SFA's input labels:

\begin{enumerate}
\item When $(p, (\text{None}, f), q)$ is a transition in $\Delta_t$, i.e. the input character is $\varepsilon$. Consequently, the SFA $\mathcal{A}$ remains in its current state, and the product transition produces the output $(f~\text{None})$.

\item When $(p, (\text{Some}~\sigma_1, f), q)$ is a transition in $\Delta_t$, synchronization is possible only with SFA transitions that share characters with $\sigma_1$, i.e., $\sigma_1\cap\sigma_2\neq \emptyset$. The resulting output is $F~f~(\sigma_1\cap\sigma_2)$, where $\sigma_2$ represents the input label of the corresponding SFA transition.
\end{enumerate}

@@

\section{An Application to String Solver}
\label{sec-app-str-solver}

>>>>

\documentclass[a4paper,UKenglish,cleveref, autoref, thm-restate]{lipics-v2021}
%This is a template for producing LIPIcs articles. 
%See lipics-v2021-authors-guidelines.pdf for further information.
%for A4 paper format use option "a4paper", for US-letter use option "letterpaper"
%for british hyphenation rules use option "UKenglish", for american hyphenation rules use option "USenglish"
%for section-numbered lemmas etc., use "numberwithinsect"
%for enabling cleveref support, use "cleveref"
%for enabling autoref support, use "autoref"
%for anonymousing the authors (e.g. for double-blind review), add "anonymous"
%for enabling thm-restate support, use "thm-restate"
%for enabling a two-column layout for the author/affilation part (only applicable for > 6 authors), use "authorcolumns"
%for producing a PDF according the PDF/A standard, add "pdfa"

@@

%TODO mandatory: add short abstract of the document
\begin{abstract}
Finite Automata (FAs) are fundamental components in the domains of programming languages. For instance, regular expressions, which are pivotal in languages such as JavaScript and Python, are frequently implemented using FAs.
%
Finite Transducers (FTs) extend the capabilities of FAs by enabling the transformation of input strings into output strings, thereby providing a more expressive framework for operations that encompass both recognition and transformation. Despite the various formalizations of FAs in proof assistants such as Coq and Isabelle/HOL, these formalizations often fall short in terms of applicability to real-world scenarios. A significant limitation of classical FAs and FTs is that transition labels are typically defined as a single character from a finite alphabet. However, in practical applications, the alphabet of an FA or FT can be enormously large or even \emph{infinite}. The classical approach to formalizing transitions can result in transition explosion, leading to critical performance bottlenecks of FA and FT operations.

@@

Finite Automata (FA) and Finite Transducers (FT) are fundamental constructs in the theory of formal languages, with extensive applications in programming languages and software engineering. For example, recent advancements in string solvers, as demonstrated by \cite{pacmpl/ChenFHHHKLRW22}, have illustrated the relationship between regular expressions in modern programming languages and various forms of FAs and FTs. Additionally, FAs and FTs find significant industrial applications, such as in the verification of AWS access control policies \cite{DBLP:conf/fmcad/BackesBCDGLRTV18}.

@@

Symbolic Finite Automata (SFA) and Symbolic Finite Transducers (SFT) \cite{cav/DAntoniV17, VeanesHLMB12Transducer} represent advanced extensions of classic FAs and  FTs, enhancing their applicability in practical scenarios.
%
These symbolic models utilize transition labels defined by first-order predicates over boolean algebras, allowing for more expressive representations. For example, a transition label might be specified as an interval ($'\texttt{a}'-'\texttt{z}'$), encompassing all characters from $'\texttt{a}'$ to $'\texttt{z}'$, or as an arithmetic condition ($x \mod 2 = 0$), denoting all even numbers. This symbolic approach not only provides a more succinct representation but also supports infinite alphabets, thereby extending the expressive capabilities of FAs and FTs.

However, the formalization of transition labels in SFAs and SFTs presents significant challenges within interactive proof assistants. Two fundamental considerations arise: first, the representation of transition labels must be sufficiently expressive to accommodate diverse predicate representations of boolean algebras; second, the formalization framework must be designed with extensibility in mind to facilitate the incorporation of new boolean algebras while minimizing redundant proof efforts.

Prior work of CertiStr \cite{cpp/KanLRS22} has successfully formalized SFAs within Isabelle/HOL, demonstrating both the efficiency and effectiveness of SFAs in practice. However, the formalization of SFTs remains an open challenge, primarily due to their inherent complexity in two aspects: the formalization of transition labels and the specification of transition output functions. SFTs constitute a significantly more expressive and powerful theoretical framework compared to  SFAs, as evidenced by their capability to model complex string transformations such as replacement operations.
%
Furthermore, numerous studies have demonstrated the broad applicability of SFTs across diverse domains. The work of \cite{VeanesHLMB12Transducer} showcases SFTs in security-critical applications, particularly for cross-site scripting (XSS) prevention. \cite{uss/HooimeijerLMSV11} extends this security focus by applying SFTs to web application sanitizer analysis. In system verification, \cite{osdi/YaseenABCL20} employs SFTs for runtime behavior monitoring, while \cite{pldi/HuD17} demonstrates their utility in automated program transformation through systematic inversion techniques.

In this work, we present a comprehensive formalization of SFTs. To address the extensibility challenges inherent in supporting diverse transition label theories, we adopt a refinement-based approach. At the abstraction level, transition labels are formalized through the fundamental mathematical concept of \emph{sets}. This abstraction facilitates subsequent refinement to various representations of Boolean algebras, such as intervals and arithmetic predicates, while maintaining theoretical consistency.

The key operation of our formalization is the product operation between an SFT and an input regular language. Specifically, given an SFA $\mathcal{A}$ representing a regular language and an SFT $\mathcal{T}$, we define the product operation $\mathcal{T} \times\mathcal{A}$ that characterizes the output language generated by $\mathcal{T}$ when processing inputs from the language recognized by $\mathcal{A}$.

In the refinement level, we implemented transition labels using an interval-based representation to examplify the refinement process. The formalization of interval algebra provides efficient set-theoretic operations—including membership checking, intersection, and difference computations—facilitating the refinement of transition labels from abstract sets to concrete intervals. Furthermore, leveraging the data refinement framework \cite{DBLP:conf/itp/Lammich13} in Isabelle/HOL, we store states and transitions using sophisticated data structures such as hashmaps and red-black trees, ensuring efficient automata manipulation.

@@

In summary, our formalization makes the following contributions:
\begin{enumerate}
\item We present the first formalization of symbolic finite transducers in Isabelle/HOL proof assistant.
\item We leverage the refinement framework to create an extensible SFT formalization that is capable of accommodates various boolean algebras, ensuring adaptability to future transition label developments.
\item We develop an interval algebra with efficient set-theoretic operations, enabling refinement of transition labels from abstract sets to concrete intervals.
\item We demonstrate the practical utility of our formalization through its application to string constraint solving, specifically in modeling replacement operations with verified correctness guarantees.
\end{enumerate}

The remainder of this paper is organized as follows:
Section \ref{sec:formalization} introduces the formalization of symbolic finite transducers.
Section \ref{sec:product-operation} presents the product operation at the abstract level.
Section \ref{sec_alg_refinement} details the algorithmic refinement of the product operation.
Section \ref{sec-app-str-solver} demonstrates the application to string constraint solving.
Section \ref{sec:related-work} discusses related work.
Section \ref{sec:conclusion} concludes with future directions.

\section{Formalization of SFTs}
\label{sec:formalization}

We begin by presenting a mathematical definition of SFTs \cite{VeanesHLMB12Transducer}, abstracting from the specific implementation details of Isabelle/HOL. 
\subsection{The Mathematical Definition of SFTs}

@@

\begin{definition}[Symbolic Finite Transducer]
\label{def-sft}
   A Symbolic Finite Transducer over $\tau_1\rightarrow \tau_2$ is a quadruple $\mathcal{T} = (\mathcal{Q}, \Delta, \mathcal{I}, \mathcal{F})$, where 
   \begin{itemize}
   \item $\mathcal{Q}$ is a finite set of states,
   \item $\mathcal{I}\subseteq \mathcal{Q}$ is the set of initial states,
   \item $\mathcal{F} \subseteq\mathcal{Q}$ is the set of accepting states,
   \item $\Delta$ is the set of transition relations. Each element in $\Delta$ is of the form $(q, \phi, f, q')$ or written as $q\xrightarrow{\phi, f} q'$, where $q$ and $q'$ are states in $\mathcal{Q}$.
   $\phi$ is a predicate of type $\tau_1\rightarrow \mathbb{B}$.
   $f$ is a lambda term of type $\tau_1\rightarrow \tau_2$. $f$ is called an \emph{output function}.
   \end{itemize}

@@

    Figure \ref{fig-example-ft} illustrates an SFT that accepts words matching the regular expression $/['\texttt{a}'-'\texttt{z}']+['\texttt{A}'-'\texttt{Z}']+/$. The transition labels utilize intervals of the form $[i\text{-}j]$, which are interpreted as predicates $\lambda x.~i \leq x \leq j$. The output functions $f_1$ and $f_2$ perform case transformations: $f_1 = \lambda x.~\texttt{toUpper}(x)$ converts lowercase letters to uppercase, while $f_2 = \lambda x.~\texttt{toLower}(x)$ performs the inverse operation. For example, given the input string \texttt{"bigSMALL"}, this SFT produces the output \texttt{"BIGsmall"}.

@@

\begin{figure}[hbt!]
	\begin{lstlisting}
record (|$'q,$||$~'a,$| |$'i,$| |$'b$|) |$\texttt{NFT}$| =
	|$\mathcal{Q}_t$| :: "|$'q$| set"
	|$\Delta_t$| :: "(|$'q,$||$~'a,$| |$'i$|) LTTS"
	|$\mathcal{I}_t$| :: "|$'q$| set"
	|$\mathcal{F}_t$| :: "|$'q$| set"
        |$\mathcal{M}_t$| :: "|$'i\Rightarrow ('a,~'b)$| Tlabel"
type_synonym |$('q,~'a,~'i)$| LTTS = "|$('q \times ('a \text{ set option }\times 'i) \times 'q)$| set"
type_synonym |$('a,~'b)$| Tlabel = "|$'a \text{ option}~\Rightarrow~'b$ set option"
	\end{lstlisting}
\caption{The formalization of $SFTs$ in Isabelle/HOL}
\label{fig-def-FT}
\end{figure}

Figure \ref{fig-def-FT} presents our formalization of SFTs in Isabelle/HOL. While the elements $\mathcal{Q}_t$, $\mathcal{I}_t$, and $\mathcal{F}_t$ directly correspond to their counterparts ($\mathcal{Q}$, $\mathcal{I}$, and $\mathcal{F}$) in Definition \ref{def-sft}, the transition relations are not exactly the same.
%
The transition relations are formalized through \texttt{LTTS} (Labeled Transducer Transition System), where each transition is represented as a triple $'q \times ('a \texttt{ set option }\times 'i) \times 'q$. This representation reflects several key design decisions aimed at enhancing the abstraction and flexibility of our SFT formalization.

Firstly, $'a \text{ set option }$ is the input type of the transition, it accepts a set of elements of type $'a$ or $\texttt{None}$ corresponding to empty string $\varepsilon$. 
Accepting a set of $'a$ elements aims to express the same but more abstract semantics of the input labels in Definition \ref{def-sft}, in which an input label is a predicate. A predicate's semantics as introduced before represents a set of elements that make the predicate true. But predicates have various different forms. For instance, the interval $[1-9]$ represents the set $\{e \mid 1 \leq e \leq 9\}$. The predicate $\lambda b.~ b[7] = 1$, where $b$ is a bit vector of length 8, denotes the set of bit vectors that have a 1 in the 7th position. All these different forms of labels are abstracted as sets in our formalization.
%
The value $\texttt{None}$ represents the empty string $\varepsilon$ in our formalization, indicating a transition that consumes no input but may still produce output elements. This design choice facilitates the modeling of real-world applications in SFTs, as we will demonstrate in our application to string solvers.

The second element of type $'i$ serves as an index into the output function space. 
%
We use indices instead of functions themselves to enable the reuse of the same output functions for different transitions.
%
The mapping $\mathcal{M}_t$ associates each index with a specific output function. These output functions, formalized by \texttt{Tlabel}, map a single input element to a set of possible output elements rather than to a single element. This design enables non-deterministic output behavior, where the transducer may select any element from the output set randomly or according to specified criteria. Additionally, output functions can produce the empty string $\varepsilon$ by returning $\texttt{None}$, providing further flexibility in transition behavior.

@@

\section{The Product Operation of SFTs}
\label{sec:product-operation}

In this section, we formalize the product operation between an SFT $\mathcal{T}$ and an SFA $\mathcal{A}$, denoted as $\mathcal{T} \times \mathcal{A}$ in Section \ref{sec:formalization}. An additional consideration in this operation is the presence of $\varepsilon$-transitions in our SFT formalization, which implies that the resulting automata may also contain $\varepsilon$-transitions. Since CertiStr \cite{cpp/KanLRS22} does not include a formalization of SFAs with $\varepsilon$-transitions, we extend the framework in CertiStr with a formalization of symbolic SFAs with $\varepsilon$-transitions (denoted as $\varepsilon$SFA) and provide a verified conversion to standard SFAs. In this paper, we present only the definitions of  $\varepsilon$SFAs and SFAs, while the complete formalization, including correctness proofs, is available in our Isabelle development.

@@

type_synonym |$('q, 'a)$| LTS = "|$'q \times 'a \text{ set }\times 'q$|"    
	\end{lstlisting}
\caption{The formalization of $\varepsilon$SFAs and SFAs}
\label{fig-def-FAs}
\end{figure}

@@

\begin{figure}[hbt!]
	\begin{lstlisting}
definition productT :: |$"('q,'a,'i,'b)"$| NFT |$\Rightarrow$| |$('q,'a)$| NFA |$\Rightarrow$| 
           |$(('a,'b)$| Tlabel |$\Rightarrow 'a \text{ set}\Rightarrow 'b\text{ set option})\Rightarrow$|
           |$('q\times 'q, ~'b)$| eNFA where
  "productT |$\mathcal{T}$| |$\mathcal{A}$| F = |$\llparenthesis$|
    |$\mathcal{Q}e = \mathcal{Q}_t~\mathcal{T} \times \mathcal{Q} ~\mathcal{A}$|,
    |$\Delta_e = \{((p,p'),~\texttt{the }(((\mathcal{M}_t~\mathcal{T})~f)\texttt{ None}),~(q, p'))\mid p, p', q, f.~p'\in \mathcal{Q}~\mathcal{A}~\land$|
          |$(p, (\texttt{None}, f), q)\in \Delta_t~\mathcal{T} \land \exists S.~(\mathcal{M}_t~\mathcal{T})~f~\texttt{None} = \text{Some } S\} ~\cup$|
        |$\{((p,p'),~\texttt{the }(F~((\mathcal{M}_t~\mathcal{T})~f)~(\sigma_1\cap\sigma_2)),~(q, q'))\mid p, p', q, \sigma_1, \sigma_2, q', f.~$|
          |$ (p, (\text{Some }\sigma_1, f), q)\in \Delta_t~\mathcal{T} \land (p', \sigma_2, q')\in\Delta~\mathcal{A}\land \sigma_1\cap \sigma_2 \neq \emptyset~\land$|
            |$ \exists S.~F~((\mathcal{M}_t~\mathcal{T})~f)~(\sigma_1\cap\sigma_2) = \text{Some } S\}$|,
    |$\Delta_e' = \{((p,p'),~(q, p'))\mid p, p', q, f.~p'\in \mathcal{Q}~\mathcal{A} \land$|
          |$(p, (None, f), q)\in \Delta_t~\mathcal{T} \land (\mathcal{M}_t~\mathcal{T})~f~\texttt{None} = \text{None}\} ~\cup$|
        |$\{((p,p'),~(q, q'))\mid p, p', q, \sigma_1, \sigma_2, q', f.~$|
          |$ (p, (\text{Some }\sigma_1, f), q)\in \Delta_t~\mathcal{T} \land (p', \sigma_2, q')\in\Delta~\mathcal{A}\land \sigma_1\cap \sigma_2 \neq \emptyset~\land$|
            |$ \exists x\in(\sigma_1\cap\sigma_2).~((\mathcal{M}_t~\mathcal{T})~f)~(\text{Some } x) = \text{None} \}$|,
    |$\mathcal{I}_e = \mathcal{I}_t~\mathcal{T} \times \mathcal{I}~\mathcal{A}$|
    |$\mathcal{F}_e = \mathcal{F}_t~\mathcal{T} \times \mathcal{F}~\mathcal{A}$| |$\rrparenthesis$|"
	\end{lstlisting}
\caption{The formalization of product operation}
\label{fig-def-FTProd}
\end{figure}

Figure \ref{fig-def-FTProd} depicts the abstract level formalization for the product of an SFT and an SFA. The parameters $\mathcal{T}$ and $\mathcal{A}$ are an SFT and an SFA, respectively. The 
result of the product operation is an $\varepsilon$SFA.
But we need to explain the role of parameter $\text{\texttt{F}}$. The output function $f$ for each transition in $\Delta_t$ is of type "$'a\;\text{option} \Rightarrow 'b\;\text{set option}$", which applies to a \emph{single} element of type $'a$ or $\varepsilon$. $\text{\texttt{F}}$ extends $f$ to apply $f$ to a set of elements. More precisely, let $f$ be an output function, the semantics of $\text{\texttt{F}}$ is defined as follows:

\[\texttt{F}~f~A=\texttt{Some } (\bigcup_{a\in A} (\text{if }f~a= \text{Some }S \texttt{ then } S \texttt{ else } \emptyset))\].

The transition relations $\Delta_e$ and $\Delta_e'$ are determinted by considering two distinct cases based on the nature of transitions in the SFT: $\varepsilon$-transitions and non-$\varepsilon$-transitions. In both cases, the transition labels in the resulting $\varepsilon$SFA are derived from the composition of the SFT's output function and the SFA's input labels. Let us consider $\Delta_e$ first.

\begin{enumerate}
\item When $(p, (\text{None}, f), q)$ is a transition in $\Delta_t$, i.e. the input character is $\varepsilon$, and $f~\text{None} \neq \text{None}$. Consequently, the SFA $\mathcal{A}$ remains in its current state, and the product transition produces the output "$((\mathcal{M}_t~\mathcal{T})~f)~\text{None}$". Remember that $f$ is just an index, $(\mathcal{M}_t~\mathcal{T})~f$ is the output function.

\item When $(p, (\text{Some}~\sigma_1, f), q)$ is a transition in $\Delta_t$, synchronization is possible only with SFA transitions that share characters with $\sigma_1$, i.e., $\sigma_1\cap\sigma_2\neq \emptyset$, where $\sigma_2$ represents the input label of the corresponding SFA transition. The resulting output is $\texttt{F}~((\mathcal{M}_t~\mathcal{T})~f)~(\sigma_1\cap\sigma_2)$.
\end{enumerate}

@@

\section{An Application to String Solving}
\label{sec-app-str-solver}
